Web-based Supplementary Materials for Combinatorial Mixture Models for Single-Cell Assays with Application to Vaccine Studies by Greg Finak, Steve De Rosa, Mario Roederer and Raphael Gottardo

### 2012-07-30 MIMOSA.tex
Derivations done. Just the MCMC for Dir-Multinomial left.
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

%\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
%\affil[2]{HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
%\affil[3]{Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}
\author{Greg Finak$^{1,*}$\email{gfinak@fhcrc.org},
SC De Rosa$^{2,**}$\email{sderosa@fhcrc.org},
Mario Roederer$^{3,***}$\email{marior@mail.nih.gov},
Raphael Gottardo$^{1,****}$\email{rgotard@fhcrc.org}\\
$^{1}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
$^{2}$HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
$^{3}$Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}

@@

\subsection{ICS}
Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, log fold-change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders. Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders). The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold-change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065}, panels A,C,E. 
At an FDR between 10-20\%, MIMOSA would lead to about 20\% more true positive being detected. Our comparison also show that ranking based on log-fold change alone is not reliable and should not be used.
In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065}, panels B,D,F). Here we presented the results based on IL2 and IFN$\gamma$ alone and the subset IL2 and/or IFN$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}. These results are consistent for other cytokines and cytokine combinations (see Web Figure A).

@@

\subsection{Polyfunctionality in Fluidigm Single-Cell Gene Expression Data}
As a proof-of-concept, we fitted our multivariate MIMOSA model looking at two specific genes in the Fluidigm data, namely BIRC3 and CCL5. In this case, $K=2$ and we have four possible combinations. 
In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the BIRC3 and CCL5 genes in unstimulated and stimulated samples (Figure~\ref{fig:polyfunctionality} A,B). Only CCL5 positive cells express BIRC3, and its expression increases upon stimulation. The typical approach to analyzing poly-functional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations  as in IL2$+$ and/or IFN$\gamma+$)\footnote{We have not discussed that yet. Perhaps we could add a sentence in the previous result section to discuss it and refer to the figure in supplementary material} would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions. That is to say, the number of BIRC3-/CCL5+ cells decreases upon stimulation and the number of BIRC3+/CCL5+ cells increases. When marginalizing over these cell populations, no difference is apparent in any of the samples. In contrast, multivariate MIMOSA tests all polyfunctional cell subpopulations simultaneously, identifying significant differences between stimulated and unstimulated conditions in 13 of the 16 samples (Figure~\ref{fig:polyfunctionality} D, black labels). Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the subject combinations, which requires multiplicity adjustment and a potential loss of power.

>>>>

%\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
%\affil[2]{HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
%\affil[3]{Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}
\author{Greg Finak$^{1}$, Andrew McDavid$^{1}$, \ldots, SC De Rosa$^{2}$, Mario Roederer$^{3}$, and Raphael Gottardo$^{1,*}$\email{rgottard@fhcrc.org}\\
$^{1}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
$^{2}$HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
$^{3}$Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}

@@

\subsection{ICS}
Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, log fold-change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders. Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders). The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold-change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065}, panels A,C,E. 
At an FDR between 10-20\%, MIMOSA would lead to about 20\% more true positive being detected. Our comparison also show that ranking based on log-fold change alone is not reliable and should not be used.
In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065}, panels B,D,F). Here we presented the results based on IL2 and IFN$\gamma$ alone and the subset IL2 and/or IFN$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}. These results are consistent for other cytokines and cytokine combinations (see Web Figure A)\footnote{Add a sentence discussing the AND/OR combination}.

@@

\subsection{Polyfunctionality in Fluidigm Single-Cell Gene Expression Data}
As a proof-of-concept, we fitted our multivariate MIMOSA model looking at two specific genes in the Fluidigm data, namely BIRC3 and CCL5. In this case, $K=2$ and we have four possible combinations. 
In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the BIRC3 and CCL5 genes in unstimulated and stimulated samples (Figure~\ref{fig:polyfunctionality} A,B). Only CCL5 positive cells express BIRC3, and its expression increases upon stimulation. The typical approach to analyzing poly-functional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations  as in IL2$+$ and/or IFN$\gamma+$) would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions. That is to say, the number of BIRC3-/CCL5+ cells decreases upon stimulation and the number of BIRC3+/CCL5+ cells increases. When marginalizing over these cell populations, no difference is apparent in any of the samples. In contrast, multivariate MIMOSA tests all polyfunctional cell subpopulations simultaneously, identifying significant differences between stimulated and unstimulated conditions in 13 of the 16 samples (Figure~\ref{fig:polyfunctionality} D, black labels). Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the subject combinations, which requires multiplicity adjustment and a potential loss of power.

### 2012-07-30 WebSupplementaryMaterialMIMOSA.tex
Bunch of updates, fixed IFNg AND OR IL2 Figure for ENV-1-PTEG
Signed-off-by: Greg Finak <gfinak@fhcrc.org>
Derivations done. Just the MCMC for Dir-Multinomial left.
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

\section*{}
\subsection{HVTN065 vaccine trial ICS data description}
\label{supp:statpublished}
HVTN065 is a phase 1 (safety and immunogenicity) trial of GeoVax HIV/AIDS DNA and MVA vaccine in 120 individuals (100 vaccinees, 20 placebo recipients, parts A and B). CD4 and CD8 T--cell epitope specific immune responses were measured via the ICS assay. 
% No need to talk about this
% Other humoral and cellular immune responses were measured via ELISA, and neutralizing antibody assays. 
Cytokines measured in the ICS assay included IFNg, TNFa, IL2, and IL4, and antigens included three Env, three Gag, and three Pol peptide pools. Results of the trial have been published in \cite{Goonetilleke:2006jk}.

@@

\subsection{Computational details for the Dirichlet-multinomial model}
\noindent\textbf{Marginal likelihood derivations}\\
Because our Dirichlet-multinomial is a direct extension of the beta-binomial model, the marginal likelihoods are obtained in the exact same fashion.The derivations is described below,

\noindent\textbf{MCMC algorithm}\\
GREG: Look at the what I did above, and do something similar here.
\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
%   \includegraphics[width=.75\columnwidth]{Figures/2}\\ 
%   \includegraphics[width=.75\columnwidth]{Figures/12} 
\begin{tikzpicture} [auto, node distance=0cm]
\node at (0,0) (A){
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) (a) {\includegraphics[width=.5\columnwidth]{Figures/15}};
    \begin{scope}[x={(a.south east)},y={(a.north west)}]
        \node at (0,1) [font=\small\sffamily] {A} ;
        \end{scope}
        \end{tikzpicture}
 };
 \node [right=of A] (B) {
 \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=.5\columnwidth]{Figures/3}};
        \begin{scope}[x={(b.south east)},y={(b.north west)}]
        \node at (0,1) [font=\small\sffamily] {B} ;
                \end{scope}
        \end{tikzpicture}
 };
 \node [below=of A] (C) {
 \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (c) {\includegraphics[width=.5\columnwidth]{Figures/5}};
        \begin{scope}[x={(c.south east)},y={(c.north west)}]
        \node at (0,1) [font=\small\sffamily] {C} ;
        \end{scope}
        \end{tikzpicture}
};
\node [below=of B] (D) {
\begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=.5\columnwidth]{Figures/6}};
        \begin{scope}[x={(d.south east)},y={(d.north west)}]
        \node at (0,1) [font=\small\sffamily] {D} ;
                \end{scope}
        \end{tikzpicture}
};
\node [below=of C] (E) {
\begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (e) {\includegraphics[width=.5\columnwidth]{Figures/7}};
        \begin{scope}[x={(d.south east)},y={(d.north west)}]
        \node at (0,1) [font=\small\sffamily] {E} ;
                \end{scope}
        \end{tikzpicture}
 };
   % \node[anchor=south west, inner sep=0] at (8,-7.5){\includegraphics[width=.5\columnwidth]{Figures/8}};
   % \node[anchor=south west, inner sep=0] at (0,-11.25){\includegraphics[width=.5\columnwidth]{Figures/9}};
    %\node[anchor=south west, inner sep=0] at (8,-11.25){\includegraphics[width=.5\columnwidth]{Figures/10}};
    %\node[anchor=south west, inner sep=0] at (0,-15){\includegraphics[width=.5\columnwidth]{Figures/11}};
    %\node[anchor=south west, inner sep=0] at (8,-15){\includegraphics[width=.5\columnwidth]{Figures/13}};

>>>>

\section*{}
\subsection{HVTN065 vaccine trial ICS data description}
\label{supp:statpublished}
HVTN065 is a phase 1 (safety and immunogenicity) trial of GeoVax HIV/AIDS DNA and MVA vaccine in 120 individuals (100 vaccinees, 20 placebo recipients, parts A and B). CD4 and CD8 T--cell epitope specific immune responses were measured via the ICS assay. 
% No need to talk about this
% Other humoral and cellular immune responses were measured via ELISA, and neutralizing antibody assays. 
Cytokines measured in the ICS assay included IFNg, TNFa, IL2, and IL4, and antigens included three Env, three Gag, and three Pol peptide pools. Results of the trial have been published in \cite{Goepfert:2011ci}.

@@

\subsection{Computational details for the Dirichlet-multinomial model}
\noindent\textbf{Marginal likelihood derivations}\\
Because our Dirichlet-multinomial is a direct extension of the beta-binomial model, the marginal likelihoods are obtained in the exact same fashion. For a non--responder, the probability of the observed data (of a single observation) is given by:

\[
\begin{split}
Pr(\mathbf{n}_s,\mathbf{n}_u|\boldsymbol{\alpha}_u)=&\int_0^1Pr(\mathbf{n}_s,\mathbf{n}_u|\boldsymbol{\alpha}_u,\mathbf{p}_u)Pr(\mathbf{p}_u|\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
=&\int_0^1Pr(\mathbf{n}_s|\boldsymbol{\alpha}_u,\mathbf{p}_u)Pr(\mathbf{n}_u|\boldsymbol{\alpha}_u,\mathbf{p}_u)Pr(\mathbf{p}_u|\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
=&\int_0^1 \mathcal{M}(\mathbf{N}_u,\mathbf{p}_u)\mathcal{M}(\mathbf{N}_s,\mathbf{p}_u)\mathrm{Dir}(\boldsymbol{\alpha}_u)\mathrm{d}\mathbf{p}_u\\
=&\int_0^1 \frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\prod_k p^{n_{sk}}_{uk} \frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\prod_k p^{n_{uk}}_{uk}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)}\prod_k p_{uk}^{\alpha_{uk}-1}\mathrm{d}\mathbf{p}_u\\
=&\frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)}\int_0^1 \prod_k p^{n_{sk}+n_{uk}+\alpha_{uk}-1}_{uk}\mathrm{d}\mathbf{p}_u\\
=&\frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\frac{\mathrm{B}(\mathbf{n}_s+\mathbf{n}_u+\boldsymbol{\alpha}_u)}{\mathrm{B}(\boldsymbol{\alpha}_u)}
\end{split}
\]
The product over all observations gives the marginal likelihood,
\[
\mathrm{L_0}(\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u)=\prod_{i=0}^P \frac{\mathbf{N}_{si}!}{\prod_k \mathbf{n}_{sik}!}\frac{\mathbf{N}_{ui}!}{\prod_k \mathbf{n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{si}+\mathbf{n}_{ui}+\boldsymbol{\alpha}_{u})}{\mathrm{B}(\boldsymbol{\alpha}_{u})}
\]

For the responder component, again the $\mathbf{n}_s$ are drawn from an independent multinomial distribution, and the $\mathbf{p}_s$ from a dirichlet with hyperparameters $\boldsymbol{\alpha_s}$, giving,
\[
\begin{split}
Pr(\mathbf{n}_s,\mathbf{n}_u|\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s)=&\int_0^1\int_0^1 \mathcal{M}(\mathbf{N}_u,\mathbf{p}_u)\mathcal{M}(\mathbf{N}_s,\mathbf{p}_s)\mathrm{Dir}(\boldsymbol{\alpha}_u)\mathrm{Dir}(\boldsymbol{\alpha}_s)\mathrm{d}\mathbf{p}_u\mathrm{d}\mathbf{p}_s\\
=&\frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\frac{1}{\mathrm{B}(\boldsymbol{\alpha}_u)\mathrm{B}(\boldsymbol{\alpha}_s)}\int_0^1\prod_k p^{n_{uk}+\alpha_{uk}-1}_{uk}\int_0^1 \prod_k p^{n_{sk}+\alpha_{sk}-1}_{sk}\mathrm{d}\mathbf{p}_u\mathrm{d}\mathbf{p}_s\\
=&\frac{\mathbf{N}_s!}{\prod_k \mathbf{n}_{sk}!}\frac{\mathbf{N}_u!}{\prod_k \mathbf{n}_{uk}!}\frac{\mathrm{B}(\mathbf{n}_u+\boldsymbol{\alpha}_u)\mathrm{B}(\mathbf{n}_s+\boldsymbol{\alpha}_s)}{\mathrm{B}(\boldsymbol{\alpha}_u)\mathrm{B}(\boldsymbol{\alpha}_s)}
\end{split}
\]
with marginal likelihood:
\[
\mathrm{L_1}(\boldsymbol{\alpha}_s,\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u)=\prod_{i=0}^P\frac{\mathbf{N}_{si}!}{\prod_k \mathbf{n}_{sik}!}\frac{\mathbf{N}_{ui}!}{\prod_k \mathbf{n}_{uik}!}\frac{\mathrm{B}(\mathbf{n}_{ui}+\boldsymbol{\alpha}_u)\mathrm{B}(\mathbf{n}_{si}+\boldsymbol{\alpha}_s)}{\mathrm{B}(\boldsymbol{\alpha}_u)\mathrm{B}(\boldsymbol{\alpha}_s)}
\]
\noindent\textbf{MCMC algorithm}\\
GREG: Look at the what I did above, and do something similar here.
\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
%   \includegraphics[width=.75\columnwidth]{Figures/2}\\ 
%   \includegraphics[width=.75\columnwidth]{Figures/12} 
\begin{tikzpicture} [auto, node distance=0cm]
 \node at (0,) (A) {
 \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (b) {\includegraphics[width=.5\columnwidth]{Figures/3}};
        \begin{scope}[x={(b.south east)},y={(b.north west)}]
        \node at (0,1) [font=\small\sffamily] {A} ;
                \end{scope}
        \end{tikzpicture}
 };
 \node [right=of A] (B) {
 \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (c) {\includegraphics[width=.5\columnwidth]{Figures/5}};
        \begin{scope}[x={(c.south east)},y={(c.north west)}]
        \node at (0,1) [font=\small\sffamily] {B} ;
        \end{scope}
        \end{tikzpicture}
};
\node [below=of A] (C) {
\begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (d) {\includegraphics[width=.5\columnwidth]{Figures/6}};
        \begin{scope}[x={(d.south east)},y={(d.north west)}]
        \node at (0,1) [font=\small\sffamily] {C} ;
                \end{scope}
        \end{tikzpicture}
};
\node [right=of C] (D) {
\begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (e) {\includegraphics[width=.5\columnwidth]{Figures/7}};
        \begin{scope}[x={(d.south east)},y={(d.north west)}]
        \node at (0,1) [font=\small\sffamily] {D} ;
                \end{scope}
        \end{tikzpicture}
 };
   % \node[anchor=south west, inner sep=0] at (8,-7.5){\includegraphics[width=.5\columnwidth]{Figures/8}};
   % \node[anchor=south west, inner sep=0] at (0,-11.25){\includegraphics[width=.5\columnwidth]{Figures/9}};
    %\node[anchor=south west, inner sep=0] at (8,-11.25){\includegraphics[width=.5\columnwidth]{Figures/10}};
    %\node[anchor=south west, inner sep=0] at (0,-15){\includegraphics[width=.5\columnwidth]{Figures/11}};
    %\node[anchor=south west, inner sep=0] at (8,-15){\includegraphics[width=.5\columnwidth]{Figures/13}};
